---
title: "Formula interface"
output: 
    md_document:
      preserve_yaml: FALSE
      fig_width: 3
      fig_height: 3
      toc: yes
      toc_depth: 2
---

# Formula interface in R

Let us take a step back and look at some data built in R. `cars` dataset
describes the speed of cars and the corresponding distance it takes to stop.

```{r}
p1 = cars %>% 
  ggplot(aes(speed, dist)) +
  geom_point()
p1
```

There is a trend in the data: with increased speed, cars stop at a longer
distance. One way to illustrate this trend comes with `ggplot2`:

```{r}
p1 + geom_smooth()
```

The result is a familiar loess function from MS Excel. The trend, however, seems unnecessary wiggly and linear trend line would probably suffice. To do that with `ggplot2`, issue `method = 'lm'` to `geom_smooth`:

```{r}
p1 + geom_smooth(method = 'lm')
```

What if we would like to know the equation for such a linear model? To do that, we use `lm` (stands for linear model) function:

```{r}
mod = lm(dist ~ speed, data = cars)
```

`lm` uses formula interface (`y ~ x`): we tell `lm` this way that we want to
explain the `distance` using `speed` and direct the function for these
variables in cars dataframe.

The intercept and slope can be printed out:

```{r}
mod
```

Formula interface is used throughout R. We can use it for plotting. Instead of:

```{r}
plot(cars$speed, cars$dist)
```

we can do:

```{r}
plot(dist ~ speed, cars)
```

Notice that in the formula interface, _y_ comes first.

It is beyond the scope, but we can add the linear model to our last plot:

```{r}
plot(dist ~ speed, cars)
abline(mod)
```
