---
title: "Growthrates package"
output: 
    md_document:
      preserve_yaml: FALSE
      fig_width: 7
      fig_height: 5
      toc: yes
      toc_depth: 2
---

# Get ready

[Growthrates](https://cran.r-project.org/web/packages/growthrates/vignettes/Introduction.html) is an R package that helps to do all sort of growth rate estimations. Broadly speaking, there are two options:

- **loglinear** -- estimate growth rate from linear part of the log
  transformed curve
- **growth model** -- assume the growth curve can be approximated by
  a specific equation


```{r}
# load the libraries
library(tidyverse)
library(growthrates)
```


# Loglinear

We will illustrate the use of `growthrates` with one of its own built-in
datasets, `bactgrowth`. The dataset contains bacterial growth measurements
upon different concentrations of tetracycline, an antibiotic. We are concerned
with the variables:

- `strain` is for three bacterial strains (`D`, `R`, `T`)
- `value` is for measured OD
- `conc` all bacterial strains were tested on across 12 concentrations (one of
  them zero i.e. no drug control) 
- `time` in hours

```{r}
str(bactgrowth)

# for more info
?bactgrowth
```

To keep it simple, let us first estimate growth rate for strain `D` at zero concentration

```{r}
sub = bactgrowth %>% 
  filter(strain=='D', conc==0)
```

The function to estimate growth rate from the linear part of log transformed
curve is `all_easylinear()`

```{r}
mod1 = all_easylinear(
  # estimate value as a function of time, 
  # grouped by replicate
  value ~ time | replicate,
  data = sub
)
```

We can see how the fit looks like by plotting

```{r}
# set the layout for two plots (two replicates)
par(mfrow=c(1, 2))
plot(mod1, log='y')
```
   
Looks good. The red points highlight the timepoints used for estimating the growth rate, default minimum value is 5 and can be adjusted by `h` parameter of `all_easylinear()` function. 

You can get the coeficients, most importantly the growth rate `mumax`, by the conventional R function `coef()` or by `results()`:

```{r}
coef(mod1)
#      y0       y0_lm     mumax      lag
# 1 0.018 0.012348195 0.2048985 1.839261
# 2 0.014 0.008513917 0.2765472 1.798446
results(mod1)
#   replicate    y0       y0_lm     mumax      lag        r2
# 1         1 0.018 0.012348195 0.2048985 1.839261 0.9832876
# 2         2 0.014 0.008513917 0.2765472 1.798446 0.9524807
```

The latter is preferred as it gives you r^2^ and coeficients are organized by
the metadata (here, by replicates).

The formula interface allows to expand the model to fit all the strains, at every concentration and replicate as easily as:


```{r}
mod2 = all_easylinear(
  value ~ time | strain + conc + replicate,
  data = bactgrowth
)
```

And all the results

```{r}
res2 = results(mod2)
nrow(res2)  # 3 strains × 12 conc × 2 reps 

# show first six
head(res2) %>% mutate_if(is.numeric, round, 3)
```

# Growth model

There are multiple models (equations) that could potentially describe the
growth curve and `growthrates` package can be used to fit several of them, including the user defined model and the four parametric log logistic model we used to fit dose-response curves. 

We are going to focus on one of the most widely used model of [Baranyi and
Roberts](https://pubmed.ncbi.nlm.nih.gov/7873331/). This one:

- requires `grow_baranyi()` function to be called within `all_growthmodels()`
  function
- needs to be provided some sane starter values for parameters, which it uses
  to start the search for final parameters
 
```{r}
# initial parameters and constraints
# y0:     starting OD
# mumax:  growthrate
# K:      final OD
# h0:     the greater it is the greater the lag phase 
p =   c(y0 = 0.02,  mumax = 1,    K = 0.1,   h0 = 1)  # intial pars start

mod3 = all_growthmodels(
  value ~ grow_baranyi(time, parms) | replicate,
  data = sub,
  p = p, 
  transform = "log"
)
```

Optionally, `all_growthmodels()` can also be provided with some lower and
upper limits for parameters, which are sometimes very useful. For instance,
one could set the upper limit of mumax to 2, which corresponds to about 20 min
doubling time. This can be done like this:

```{r}
p   = c(y0 = 0.02,  mumax = 1,    K = 0.1,  h0 = 1)  # intial pars start
lwr = c(y0 = 0.01,  mumax = 1e-2, K = 0.06, h0 = 0)  # lower limits
upr = c(y0 = 0.06,  mumax = 2,    K = 1,    h0 = 30) # upper limits

mod4 = all_growthmodels(
  value ~ grow_baranyi(time, parms) | replicate,
  data = sub,
  p = p, lower = lwr, upper = upr,
  transform = "log"
)
```

One should not take it casually which initial parameters are provided. If the starter values are too low, for instance, it is rather easy for the algorithm to arrive at some local minimum. Some trial-and-error here goes a long way: fitting and plotting (preferably many different curves in parallel) to see what works. Don't expect an ideal solution though, find out what works the best and move on.

How do the resulting fits look like?

```{r}
par(mfrow=c(1, 2))
plot(mod4, log='y', ylim=c(0.01, 0.64))
```

For this particular data, the Baranyi & Roberts model is not the best fit (or if you find better starting parametes & limits, let me know vvarik.mail@gmail.com). It can prove to be very useful, however, if there are not enough datapoints on the linear part of the exponential growth phase.

And the parameters can be extracted as before:

```{r}
res4 = results(mod4)
res4 %>% mutate_if(is.numeric, round, 3)
```
